---
title: "How Bike Theft Trends in Toronto Have Changed Over Time"
subtitle: "And How Cyclists Can Keep Their Bikes Safe"
author: "Ricky Yuen, Xinyi Zhang"
date: "3/5/2021"
output: pdf_document
---

```{r setup, include=FALSE}
# don't show code in data-driven article
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r}
# set up libraries and load data
library(ggmap)
library(ggplot2)
library(tidyverse)
library(latticeExtra)

data <- read.csv("bicycle-thefts-2014-2019.csv")
```

```{r}
data_recovered <- data %>% group_by(Status) %>%
  summarise(count=n())
```


# Introduction
In August of 2020, a group of bike thieves made headlines when they were caught on video attempting to steal several bikes from a gated garage in downtown Toronto using a power saw. 11 months earlier, a thief was filmed cutting off the lock of a $1500 bike in Toronto's financial district in broad daylight. Unfortunately, videos like these capture only a small proportion of the bike thefts that occur in Toronto every year. According to the bike theft reports released by the Toronto Police Service, over 21,500 bikes have been stolen from 2014 to 2019. Unfortunately, only around 1% of those bikes have been recovered. 

In this article, we will analyze the aforementioned bike theft reports to explore how bike theft trends have changed over time. Cyclists can leverage the patterns we have identified to assess and minimize their risk of becoming a victim of bike theft.

# How have bike theft numbers changed over the past few years at the monthly level?
```{r}
data$Cost_of_Bike[is.na(data$Cost_of_Bike)] <- 0
data_month <- data %>%
  group_by(Occurrence_Year, Occurrence_Month) %>%
  summarise(frequency=n(), cost=sum(Cost_of_Bike))
data_month$Occurrence_Date <- zoo::as.yearmon(paste(data_month$Occurrence_Year, data_month$Occurrence_Month), "%Y %m")
months <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12)
temps <- c(-3.5, -3, -1.5, 7.5, 16.5, 19.5, 23, 22, 18, 23, 5, 0)
data_month <- data_month %>% mutate(Temperature=plyr::mapvalues(Occurrence_Month, months, temps))
```

```{r}
ggplot(data=data_month) + 
  geom_line(aes(x=Occurrence_Date, y=frequency, color=Temperature)) +
  geom_point(aes(x=Occurrence_Date, y=frequency, color=Temperature)) + 
  scale_color_gradient(low="blue4", high="red3") +
  labs(x="Date",
       y="Number of Thefts") + 
  theme_minimal()
```
After grouping bike thefts by month, the most noticeable pattern that emerges is the cyclic trend between the seasons and theft counts. In the winter when temperatures often fall below freezing, bike theft counts drop to their lowest counts (generally, falling below 100 thefts). On the other hand, in the summer, bike theft counts reach their peak with typically over 500 thefts in the warmest months of the year. 

This seasonal pattern makes sense considering that cyclists are more likely to regularly bike when the weather is warm. This in turn affords bike thieves more opportunities to spot and steal bikes during the summer months compared to the winter months. In fact, our analysis indicates that over 5 times the number of bikes stolen in the winter will be stolen in the summer of that year. Therefore, cyclists should strive to be extra vigilant when storing and locking up their bikes in the summer.

We can also observe that as the years progress, the number of thefts appear to increase during both the summer and winter months. Therefore, we'd like to determine how bike thefts numbers in Toronto have changed over time when weather isn't a factor. 

# How have bike theft numbers changed at the yearly level?
```{r}
data_year <- data_month %>%
  group_by(Occurrence_Year) %>%
  summarise(frequency=sum(frequency),
            cost=sum(cost)) %>%
  mutate(cost = cost/1000000)

obj1 <- xyplot(frequency ~ Occurrence_Year, data_year, type="l", xlab="Year", ylab="Number of Thefts")
obj2 <- xyplot(cost ~ Occurrence_Year, data_year, type="l", 
               xlab="Year", ylab="Cost (in Millions)",
               col="#00BA38")
update(doubleYScale(obj1, obj2, add.ylab2 = TRUE),
  par.settings = simpleTheme(col = c('#619CFF','#00BA38')))
```
By aggregating bike theft counts over the course of each year (displayed in blue), we can focus on how the frequency of this crime has changed since 2014 ignoring seasonal factors. From our visualization, we notice that the number of bike thefts has steadily increased for each year between 2014 and 2018. The largest increase occurred between 2015 and 2016 which saw 15% jump in stolen bikes. However, this trend was broken in 2019 which saw a decrease in the number of bikes thefts. 

Interestingly, although the number of bike thefts decreased in 2019, the profit (in green) made by bike thieves did not. In fact, although the number of bike thefts decreased by a count of 300 between 2018 and 2019, the cost of the bikes stolen in 2019 totaled over $100,000 more than the cost of the bikes stolen in 2018. This indicates that bike thieves have become more discerning about the cost and potential resale value of the bikes they steal. Therefore, amateur cyclists are advised to avoid purchasing extravagant bicycles because they may be easily targeted by bike thieves.

Considering that bike theft counts have generally increased over time, we'd like to determine the safest place for cyclists to keep their bikes...by identifying where bikes have been usually stolen from over the past few years.

# How have bike theft location types changed over time?
```{r}
data_premise <- data %>%
  filter(Premise_Type != "Other") %>%
  mutate(Premise_Type=str_replace(Premise_Type, "Apartment", "Residential")) %>%
  mutate(Premise_Type=str_replace(Premise_Type, "House", "Residential")) %>%
  mutate(Premise_Type=fct_relevel(Premise_Type, "Outside", "Commercial", "Residential"))

data_premise <- data_premise %>%
  group_by(Occurrence_Year, Premise_Type) %>%
  summarise(frequency=n())

data_premise <- data_premise %>%
  group_by(Occurrence_Year) %>%
  mutate(total=sum(frequency))

data_premise <- data_premise %>%
  mutate(percentage=frequency/total)

data_premise %>% ggplot(aes(x=Occurrence_Year, y=percentage, fill=Premise_Type)) +
  geom_bar(stat="identity") +
  labs(x="Year",
       y="Percentage",
       fill="Theft Location Type") + 
  theme_minimal()
```
Two main patterns stand out when we look at the change in theft location type over time: the percentage of thefts occurring at a cyclist's place of residence has increased over time while the percentage of thefts occurring outside has decreased over time. More specifically, the percentage of thefts occurring outside has decreased from 49% to 35% while the percentage of thefts occurring at a residence has increased from 39% to 52%. Therefore, although a plurality of thefts took place outside in 2014, a majority of thefts now take place at an apartment or house now.

We can speculate that due to a more vigilant public, over time thieves find it increasingly difficult to steal bikes in broad daylight. For example, sawing open an expensive bike lock is a lot more suspicious than discreetly picking open a cheap lock. Therefore, thieves seem to have shifted tactics and are now targeting residential areas where they are less likely to be caught. Therefore, it looks like cyclists should consider beefing up their home security, perhaps taking advantage of motion sensor alarms to ward off bike thieves.

To determine if our speculation is correct, let's find the time of day when bike thefts typically occur. If the majority of bike thefts occur during the night, when cyclists have stored their bikes at home, our hypothesis would be supported.

# At what time during the day do bike thefts typically occur? 
```{r}
data_hourly <- data %>% mutate(Hour = lubridate::hour(Occurrence_Time)) %>%
  group_by(Occurrence_Year, Hour) %>%
  mutate(frequency=n())

data_hourly %>% ggplot(aes(x=Hour, y=frequency)) +
  geom_line(colour="#619CFF") +
  labs(y="Number of Thefts") + 
  theme_minimal() + 
  facet_wrap(~Occurrence_Year)
```
Overall, it appears that for every year, there is a consistent bike theft pattern that changes depending on the time of day. That is, the number of bike thefts generally starts fairly high after midnight, dips to almost 0 during the morning commute, rises after cyclists park their bikes at work, and dips again during the lunch hour. Then, the number rises again after lunch, dips again during the commute home, and rises again as the night grows later. Furthermore, as the years pass, the number of thefts in the evening (when bikes are stored "safely" at home) has clearly increased. This supports our hypothesis that over time, bike thieves have focused their attention on targeting residential areas after dark. This adds credibility to our suggestion of heightening home security systems to prevent bike thefts in residential areas.

# Where in Toronto do the majority of bike thefts occur?
```{r, message=FALSE}
data <- read.csv("bicycle-thefts-2014-2019.csv")

map <- qmplot(Long, Lat, data=data, maptype="toner", geom="blank", 
              xlab="Longitude", ylab="Latitude")

map + geom_point(alpha=0.2, colour="#481567FF") + 
  stat_density2d(aes(fill=..level..), geom="polygon", alpha=.5) +
  scale_fill_viridis_c() + 
  facet_wrap(~Occurrence_Year) +
  theme(legend.position = "none")
```

This visualization shows the density of bike thefts across the city. We can observe that the frequency of bike thefts is dense around the downtown area, especially by the waterfront. Whereas the location of bike thefts are more scattered in the suburban areas of Toronto. This trend has not changed within these years, and will possibly be the trend for the years to come.

We hypothesize this finding comes from the fact there are more people riding bikes in this downtown area. From the earlier visualization, we also saw an increase in residential bike thefts. So it is notable that the population in this area is significantly more dense when compared to the upper suburban area. 

We see that bike theft is happening all over Toronto. These density plots can be useful in finding the "hotspot" for bike theft locations. So what can we do to avoid your bikes from being stolen?

# Summary

- summarize main finding and actionable insights



# References
https://data.torontopolice.on.ca/datasets/bicycle-thefts

https://toronto.ctvnews.ca/my-jaw-dropped-attempted-bike-theft-by-power-saw-wielding-suspects-caught-on-camera-1.5080725

https://www.cbc.ca/news/canada/toronto/brazen-daylight-bike-theft-1.5272602

https://www.currentresults.com/Weather/Canada/Ontario/Places/toronto-temperatures-by-month-average.php
