---
title: "How Bike Theft Trends in Toronto Have Changed Over Time"
subtitle: "And How Cyclists Can Leverage This Information to Keep Their Bikes Safe"
author: "Ricky Yuen, Xinyi Zhang"
date: "3/5/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

- bike purchase
- risk assessment
- if and how they changed over time
- bike features and risk factors associated with bike theft
story
- meaningful title
- introduction: goal
- clear sections with meaningful section headers
- for each visualization: what question the visualization addresses
  - what the reader should see that is useful/interesting for understanding the story
  - how its analysis leads to new question and the visualization that follows
- summarize main finding and actionable insights

```{r}
library(ggmap)
library(ggplot2)
library(tidyverse)
library(latticeExtra)
data <- read.csv("bicycle-thefts-2014-2019.csv")
```
- trends over months and years
- costs over time

```{r}
data$Cost_of_Bike[is.na(data$Cost_of_Bike)] <- 0
data_month <- data %>%
  group_by(Occurrence_Year, Occurrence_Month) %>%
  summarise(frequency=n(), cost=sum(Cost_of_Bike))
data_month$Occurrence_Date <- zoo::as.yearmon(paste(data_month$Occurrence_Year, data_month$Occurrence_Month), "%Y %m")
```

```{r}
ggplot(data=data_month) + 
  geom_line(aes(x=Occurrence_Date, y=frequency)) +
  geom_point(aes(x=Occurrence_Date, y=frequency))
```
```{r}
data_year <- data_month %>%
  group_by(Occurrence_Year) %>%
  summarise(frequency=sum(frequency),
            cost=sum(cost))

obj1 <- xyplot(frequency ~ Occurrence_Year, data_year, type="l")
obj2 <- xyplot(cost ~ Occurrence_Year, data_year, type="l")
doubleYScale(obj1, obj2, text=c("Frequency", "Cost"), add.ylab2 = TRUE)
```
- primary offense over time
- make, speed, color
- location of theft
- location of most thefts
- time of day
```{r}
data_offense <- data %>%
  filter(str_detect(Primary_Offence, "THEFT") | str_detect(Primary_Offence, "B&E")) %>%
  mutate(Primary_Offence=str_replace(Primary_Offence, "THEFT.*", "Theft")) %>%
  mutate(Primary_Offence=str_replace(Primary_Offence, "B&E.*", "B&E"))

data_offense <- data_offense %>%
  group_by(Occurrence_Year, Primary_Offence) %>%
  summarise(frequency=n())

data_offense %>% ggplot(aes(x=Occurrence_Year, y=frequency, fill=Primary_Offence)) +
  geom_bar(stat="identity")
```
