---
title: "How Bike Theft Trends in Toronto Have Changed Over Time"
subtitle: "And How Cyclists Can Leverage This Information to Keep Their Bikes Safe"
author: "Ricky Yuen, Xinyi Zhang"
date: "3/5/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

- bike purchase
- risk assessment
- if and how they changed over time
- bike features and risk factors associated with bike theft
story
- meaningful title
- introduction: goal
- clear sections with meaningful section headers
- for each visualization: what question the visualization addresses
  - what the reader should see that is useful/interesting for understanding the story
  - how its analysis leads to new question and the visualization that follows
- summarize main finding and actionable insights

```{r}
library(ggmap)
library(ggplot2)
library(tidyverse)
library(latticeExtra)
data <- read.csv("bicycle-thefts-2014-2019.csv")
```
- trends over months and years
- costs over time

```{r}
data$Cost_of_Bike[is.na(data$Cost_of_Bike)] <- 0
data_month <- data %>%
  group_by(Occurrence_Year, Occurrence_Month) %>%
  summarise(frequency=n(), cost=sum(Cost_of_Bike))
data_month$Occurrence_Date <- zoo::as.yearmon(paste(data_month$Occurrence_Year, data_month$Occurrence_Month), "%Y %m")
```

```{r}
ggplot(data=data_month) + 
  geom_line(aes(x=Occurrence_Date, y=frequency)) +
  geom_point(aes(x=Occurrence_Date, y=frequency))
```
```{r}
data_year <- data_month %>%
  group_by(Occurrence_Year) %>%
  summarise(frequency=sum(frequency),
            cost=sum(cost))

obj1 <- xyplot(frequency ~ Occurrence_Year, data_year, type="l")
obj2 <- xyplot(cost ~ Occurrence_Year, data_year, type="l")
doubleYScale(obj1, obj2, text=c("Frequency", "Cost"), add.ylab2 = TRUE)
```
- primary offense over time
- make, speed, color
- location of theft
- location of most thefts
- time of day
```{r}
data_offense <- data %>%
  filter(str_detect(Primary_Offence, "THEFT") | str_detect(Primary_Offence, "B&E")) %>%
  mutate(Primary_Offence=str_replace(Primary_Offence, "THEFT.*", "Theft")) %>%
  mutate(Primary_Offence=str_replace(Primary_Offence, "B&E.*", "B&E"))

data_offense <- data_offense %>%
  group_by(Occurrence_Year, Primary_Offence) %>%
  summarise(frequency=n())

data_offense %>% ggplot(aes(x=Occurrence_Year, y=frequency, fill=Primary_Offence)) +
  geom_bar(stat="identity")
```
- density map of stolen bikes
- location of thefts

```{r, message=FALSE}
data <- read.csv("bicycle-thefts-2014-2019.csv")

map <- qmplot(Long, Lat, data=data, maptype="toner", geom="blank", 
              xlab="Longitude", ylab="Latitude")

map + geom_point() +
  stat_density2d(aes(fill=..level..), geom="polygon", alpha=.2, colour="red") +
  scale_fill_gradient2(low="white", mid="yellow", high="red")

```
- most stolen bike make
```{r}
makes <- plyr::count(data, "Bike_Make")
sorted_makes <- makes[order(-makes$freq),]
sorted_makes$Bike_Make <- gsub("^OT\\s", "OTHER", sorted_makes$Bike_Make)
sorted_makes$Bike_Make <- gsub("^UK\\s", "UNKNOWN MAKE", sorted_makes$Bike_Make)
sorted_makes$Bike_Make <- gsub("^TR\\s", "TREK", sorted_makes$Bike_Make)
sorted_makes$Bike_Make <- gsub("^NO\\s", "NORCO", sorted_makes$Bike_Make)
sorted_makes$Bike_Make <- gsub("^CC\\s", "CCM", sorted_makes$Bike_Make)
sorted_makes$Bike_Make <- gsub("^GI\\s", "GIANT", sorted_makes$Bike_Make)
sorted_makes$Bike_Make <- gsub("^SC\\s", "SCHWINN", sorted_makes$Bike_Make)
sorted_makes$Bike_Make <- gsub("^CA\\s", "CANNONDALE", sorted_makes$Bike_Make)
sorted_makes$Bike_Make <- gsub("^KO\\s", "KONA", sorted_makes$Bike_Make)
sorted_makes$Bike_Make <- gsub("^RA\\s", "RALEIGH", sorted_makes$Bike_Make)
sorted_makes$Bike_Make <- gsub("^SU\\s", "SUPERCYCLE", sorted_makes$Bike_Make)
sorted_makes$Bike_Make <- gsub("\\s*$", "", sorted_makes$Bike_Make)

sorted_makes <- plyr::count(sorted_makes, "Bike_Make")
sorted_makes <- sorted_makes[order(-sorted_makes$freq),]
sorted_makes <- sorted_makes[3:12,]

ggplot(sorted_makes, aes(Bike_Make, freq)) + geom_bar(stat="identity") +
  xlab("Bike Make") + ylab("Number of Bikes Stolen") +
  scale_x_discrete(guide = guide_axis(n.dodge=2)) + theme_bw()
```
- Which neighbourhood had the most bikes stolen from

```{r}
neighbourhood <- plyr::count(data, "Neighbourhood")
sorted_neighbourhood <- neighbourhood[order(-neighbourhood$freq),]
ggplot(sorted_neighbourhood[1:10,], aes(Neighbourhood, freq)) + 
  geom_bar(stat="identity") + theme(axis.text.x = element_text(angle = 15)) + 
  ylab("Number of Bikes Stolen")
```
- Month and Year vs Number of Bikes stolen

```{r}
times <- plyr::count(data, "Occurrence_Date")

times$Occurrence_Date <- gsub("\\s.*$", "", times$Occurrence_Date) 

dates <- c(sprintf("2014/0%d/01",seq(01:12)),
           sprintf("2015/0%d/01",seq(01:12)),
           sprintf("2016/0%d/01",seq(01:12)),
           sprintf("2017/0%d/01",seq(01:12)),
           sprintf("2018/0%d/01",seq(01:12)),
           sprintf("2019/0%d/01",seq(01:12)))
dates <- gsub("/010/", "/10/", dates)
dates <- gsub("/011/", "/11/", dates)
dates <- gsub("/012/", "/12/", dates)

months <- c("Jan","Feb","Mar",
            "Apr","May","Jun",
            "Jul","Aug","Sep",
            "Oct","Nov","Dec")

date_lab <- c(months, months, months, months, months, months)

ggplot(times, aes(Occurrence_Date, freq)) + geom_point() + 
  scale_x_discrete(breaks=dates, labels=date_lab) + theme_bw() +
theme(axis.text.x = element_text(angle = 90)) + xlab("Month") + ylab("Number of Stolen Bikes") + 
  annotate("text", x = "2014/07/01", y = 0, label = "2014") + 
  annotate("text", x = "2015/07/01", y = 0, label = "2015") + 
  annotate("text", x = "2016/07/01", y = 0, label = "2016") + 
  annotate("text", x = "2017/07/01", y = 0, label = "2017") + 
  annotate("text", x = "2018/07/01", y = 0, label = "2018") + 
  annotate("text", x = "2019/07/01", y = 0, label = "2019")
```